DRAFT: The algebraic hierarchy in type theory using type classes

### 2010-02-26 jar.tex
Major rewrite.

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

We abstract from concrete algebra and formalize many-sorted universal algebra, including a
construction of
free $\Sigma$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
In fact, our aim is to use the present development as a new basis for Bishop's program in type
theory~\cite{typesreal-article}.\marginpar{More}

@@

\subsection{Type classes}
Implementation in Ltac, Notation etc
Coercions

@@

We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
\begin{code}
Record Monoid (A: Type) (R: A → A → Prop)
  (e: Equivalence R) : Type := ...
\end{code}
However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
equivalence, we end up with two non-convertible ways to express that the relation (together with
some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
interchangable, and hope for automation, such as type class instance resolution, to use them
interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
essentially elevating them to `values', which goes against the principle of proof
irrelevance.\marginpar{What if we add proof irrelevance?}

Neither of these options are very appealing. We really should just make |Monoid| take as parameters
only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
how things ought to be.

Another benefit of this approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, expressing as a type class constraint that some collection of relations and
operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as simple type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

We remark that leaving the equivalence proofs bundled means that when one has a proof that something
is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
equivalence. Fortunately, having redundant proofs like this is entirely harmless when one's terms
never refer (directly or indirectly) to proofs, which is the case when the relations and operations
they are composed of are left unbundled. Here, too, we enjoy the warm embrace of proof irrelevance.

From the discussion so far, we derive some principles:
\begin{enumerate}
 \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
(i.e. carriers, relations, and operations) as parameters, and nothing else;
 \item (as a consequence) these concepts should be defined as predicates (which includes records and
classes living in Prop), and their proofs may be kept opaque;
 \item (as a second consequence) algebraic expressions need/should never refer (directly or
indirectly) to proofs.
\end{enumerate}
% \setcounter{enumi}{3}

@@

Suppose we have defined |SemiGroup| in line with the aforementioned principles:
\begin{code}
Record SemiGroup (A: Type) (eq: A → A → Prop)
 (op: A → A → A): Prop := ...
\end{code}
Then our starting point is the obvious but awful:
\begin{code}
Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
                   eq (op (op x y) z) (op (op z y) x).
\end{code}
As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
and |op| are just local names and there is no way to predeclare that they are to be associated with
any notation. What we really need are canonical names for these notions, so that we may fashion
notations for them.

@@

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
equiv `projection' is the identity function whose argument is inferred using instance resolution.
Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
it, notations:
\begin{code}
  Record SemiGroup (A: Type) (eq: Equiv A)
                             (op: SemiGroupOp A): Prop := ...
  Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
                    x & y & z = z & y & x.
\end{code}
We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation |scope| mechanism.

It is important to point out that although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane.

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for |assoc1| looks like:
\begin{code}
  A: Type
  eq: Equiv A
  op: SemiGroupOp A
  P: SemiGroup A eq op
\end{code}

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like |eq| and |op| for |assoc1|) \emph{is} a bit of a chore. Fortunately, Coq
provides a feature called implicit generalization, which is exactly what we need.
\marginpar{describe how it works} With it, we can write |assoc1| as follows:
\begin{code}
  Lemma assoc1 `(SemiGroup A) x y z: x & y & z = z & y & x.
\end{code}
Thus, we have reached the mathematical ideal we aimed for.

@@

\section{Subset types}
BEGIN SPECULATIVE

A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
squash operator, |[]:Type->Prop|, we may define a subset type as |\{A\mid B\}=Sig A. [B]|. This is
pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
present Coq implementation. Proof irrelevanve in Coq would allow a faithful
implementation.

Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
\ldots.

Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
seems to be no reason to connect this to |Prop| or impredicativity.

END SPECULATIVE

We add another principle:
\begin{itemize}\setcounter{enumi}{3}
 \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
rewrite this subset |\{A\mid B\}| as |Record r (a:A):={b:B}|.
\end{itemize}

\section{Algebraic hierarchy using type classes}\label{classes}
\paragraph{Hierarchy.}\label{hierarchy}
In our formalization of algebraic structures we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
where e.g.\ |RingPlus| is the operation class for the semigroup operator.
The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
provide a ring.

\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.
% A similar style should be possible for, say, the |Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find
% it.
We note that |op| does not depend on the proof that |e| is an equivalence. We use Coq's implicit
quantification (|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem
and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
theorem. Both features are new in Coq and stem from the type class implementation.

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively
harmless\footnote{Coq's data structure for contexts is not very efficient however, we believe this
may be a bottleneck for huge developments. It seems that the data structure chosen
in~\cite{asperti2009compact} will behave better.}%
: most
terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

@@

Manifest fields in records allow us to fix a field of a dependent record.
Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes would
be awkward due to problems with equality. We avoided these problems by shifting the fields
from the dependent record to its argument, thus mimicking manifest fields by using type classes.
Since |op| is \emph{instantiated} with |mult|, we need not worry about the
equality |op=mult|.

\section{Universal algebra}\label{univ}
After having implemented concrete algebraic structures, we generalized some of our constructions
by implementing many-sorted universal algebra. A formalization of universal algebra
already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. It has recently been extended with more
categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of this
using type classes. We defined equational theories and their varieties:
the categories of their models. A quote function helps to translate a statement for an instance of
a type class to a statement in the equational logic.

\subsection{Category theory}\label{cats}
Using type classes, we provide a basic library for category theory: categories, functors,
adjunctions, monads,\ldots We follow the usual type theoretical definition of a\marginpar{Use
notation for comp} category~\cite{saibi1995constructive}
\begin{code}
 Class Category O `{!Arrows O} `{forall x y: O, Equiv (x --> y)}
   `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> forall x y: O, Setoid (x --> y)
  ; comp_proper:> forall x y z: O,
      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
  ; comp_assoc: forall (w x y z: O) (a: w --> x) (b: x --> y) (c: y --> z),
      comp c (comp b a) == comp (comp c b) a
  ; id_l: forall (x y: O) (a: y --> x), comp cat_id a == a
  ; id_r: forall (x y: O) (a: x --> y), comp a cat_id == a }.
\end{code}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

%format ◎ = "\circ"
For instance natural transformations are formalized as: |◎|
\begin{code}
  Context `{Category C} `{Category D} (F G: C -> D)
     `{!Functor F Fa} `{!Functor G Ga}.

  Class NaturalTransformation (η: forall c: C, F c --> G c): Prop :=
    natural: forall (x y: C) (f: x --> y), η y ◎ fmap F f == fmap G f ◎ η x.
\end{code}

The notation |◎| denotes function composition.
Here |fmap F| explicitly refers to the arrow part of the functor |F|.\\
In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
not seem to fit with type theory, hence we follow the customs in functional programming.

@@

We provide the category of setoids, the category of categories, the dual of a category, products,
and hence co-products, the category of $\Sigma$-algebras.

We provide the construction of the term-algebra for a signature $\Sigma$ and show that it is
initial in the category of $\Sigma$-algebras. We proceed to prove that this provides a left-adjoint
to the forgetful functor. The resulting expression monad will be used below.

\subsection{Universal algebra continued}\label{univ2}

We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
theorem.~\marginpar{Description of the work.}
%perhaps the soundness theorem for equational logic.

In order to prove the first homophism theorem, we define a congruence:
%theory/ua_congruence.v
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
We note the use of |Proper|.

We define the forgetful functor from the category of $\Sigma$-algebras to the category of sets.

We define the term algebra consisting of closed terms, terms with the empty type as set of
variables. % extensionally equal closed term???
We show that the term algebra is initial. One may expect that this could be useful to
\emph{define}, say, the natural numbers as the initial semiring. However, it seems difficult to
prove decidability of equality, as it seems to require a normalization procedure.

We define the product of two algebras and show that it is the categorical product.

We connect the abstract theory to the concrete theory. Concretely, let $\Sigma$ be the theory
of monoids. Given a monoid, we can construct the corresponding $\Sigma$-algebra and an
object in the category of $\Sigma$-algebras. Conversely, given an object in the category, we
construct a monoid.
\begin{code}
 Variable o: Variety theory.
 Global Instance from_object: Monoid (o tt).
\end{code}
Similarly, for arrows.

%There is a slight problem when defining the forgetful functor from the category of
%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
%categories\footnote{Perhaps we should have used higher categories for this.}.
%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
%definitions.
\section{Interfaces using category theory}\label{interfaces}\label{modul}
The current work was stimulated by our desire~\cite{Riemann} to move between several
implementations in a flexible way. In haskell type class polymorphism is used for this purpose.
Hence, we do the same in type theory.

We have characterized the naturals as the initial object in the category of semirings and derived
many of their properties from this interface. Unary, binary and machine numbers are
instances of this interface, so we can directly apply these results to all these instances.
Similarly, for the integers are the initial ring. The rationals are the field of fractions of the
integers. 
Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra of a set $X$.
We provide two implementations of polynomials: the
standard representation using lists of coefficients and the Bernstein representation. A Bernstein
basis polynomial is one of the form:
\[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.

@@

The context grows as follows:\marginpar{expand}

@@

We define a term language.
\begin{code}
Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
\end{code}
This defines a monad on Type\marginpar{Do this!}.
The expression type is parameterized over the set of variable indices. Hence, we diverge
 from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
 dummy variables.

\noindent An expression is only meaningful in the context of a variable assignment:\footnote{nat is
an Expr-algebra. eval is map vars composed with the algebra map.}
\begin{code}
Definition Value := nat.
Definition Vars V := V -> Value.

Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
  match e with
  | Zero => 0
  | Mult a b => eval vs a * eval vs b
  | Var v => vs v
  end.
\end{code}
%
Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
$\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
corresponding bind operation.

@@

We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
with extensional functions between them~\marginpar{Really?}.

\noindent Some simple combinators for variable packs:
%
\begin{code}
Definition novars: Vars False := False_rect _.
Definition singlevar (x: Value): Vars unit := fun _ => x.
Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
  fun i => if i then a j else b j.
\end{code}

>>>>

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
\usepackage{amsmath}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{DRAFT: The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

We abstract from concrete algebra and formalize many-sorted universal algebra, including a
construction of
free $\tau$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
In fact, our aim is to use the present development as a new basis for Bishop's program in type
theory~\cite{typesreal-article}.\marginpar{More}

@@

Future versions of Coq are planned to support proof irrelevance: every |p,q:P:Prop| are
convertible; see section~\cite{explit}.

\subsection{Type classes}
Implementation in Ltac, Notation etc\\
Coercions\\
Unification hints

@@

% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
% \begin{code}
% Record Monoid (A: Type) (R: A → A → Prop)
%   (e: Equivalence R) : Type := ...
% \end{code}
% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
% equivalence, we end up with two non-convertible ways to express that the relation (together with
% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
% interchangable, and hope for automation, such as type class instance resolution, to use them
% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
% essentially elevating them to `values', which goes against the principle of proof
% irrelevance.\marginpar{What if we add proof irrelevance?}
% BOTTOMLINE: Bundle as much as we can!
% 
% Neither of these options are very appealing. We really should just make |Monoid| take as
% parameters
% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
% this
% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
% how things ought to be.

Another benefit of our approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, to express that some collection of relations and
operations posess the structure as a type class constraint seems to require manifest fields; see
Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

Leaving the equivalence proofs bundled means that when one has a proof that a structure
is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
equivalence. Fortunately, this redundancy is entirely harmless since our terms
never refer (directly or indirectly) to proofs, which is the case when the relations and operations
they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
irrelevance.

From the discussion so far, we derive some principles:
\begin{enumerate}
 \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
(i.e. carriers, relations, and operations) as parameters, and nothing else;
 \item (as a consequence) these concepts should be defined as predicates (which includes records and
classes living in Prop), and their proofs may be kept opaque;
 \item (as a second consequence) algebraic expressions need/should never refer (directly or
indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
famously does refer to a proof that the element is non-zero, but not to a proof that the operations
indeed form a field.}
\end{enumerate}
% \setcounter{enumi}{3}

@@

Suppose we have defined |SemiGroup| in line with the aforementioned principles:
\begin{code}
Record SemiGroup (A: Type) (eq: A → A → Prop)
 (op: A → A → A): Prop := ...
\end{code}
Then our starting point is the obvious but awful:
\begin{code}
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                   eq (op (op x y) z) (op (op z y) x).
\end{code}
As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
and |op| are just local names and there is no way to predeclare that they are to be associated with
any notation. What we really need are canonical names for these notions, so that we may fashion
notations for them.

@@

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
equiv `projection' is the identity function whose argument is inferred using instance resolution.
Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
it, notations:
\begin{code}
Record SemiGroup (A: Type) (eq: Equiv A)
                 (op: SemiGroupOp A): Prop := ...
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                 x & y & z = z & y & x.
\end{code}
We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation |scope| mechanism.

It is important to point out that although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane; see Section~\ref{canonical}.

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for |perm| looks like:
\begin{code}
  A: Type
  eq: Equiv A
  op: SemiGroupOp A
  P: SemiGroup A eq op
\end{code}

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like |eq| and |op| for |perm|) \emph{is} a bit of a chore. Fortunately, Coq
provides a feature called implicit generalization, which is exactly what we need.
With it, we can write |perm| as follows:
\begin{code}
  Lemma perm `(SemiGroup A) x y z: x & y & z = z & y & x.
\end{code}
Thus, we have reached the mathematical ideal we aimed for.

@@

\section{Putting it all together}\label{classes}
Combining the ideas in the previous sections we end up with a formalization of algebraic
structures where we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
where e.g.\ |RingPlus| is the operation class for the semigroup operator.
The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
provide a ring.

\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.
% A similar style should be possible for, say, the |Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find
% it.
We note that |op| does not depend on the proof that |e| is an equivalence. As explained above we
use Coq's implicit quantification (|`{}|) to avoid having to write all the parameters when
\emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when
\emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively
harmless\footnote{Coq's data structure for contexts is not very efficient. Gonthier fears that this
may be a bottleneck for huge developments. It seems that the data structure chosen
in~\cite{asperti2009compact} will behave better.}%
: most terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

@@

Manifest fields in records allow us to fix a field of a dependent record. For example, as in the
example above, |Monoid with op=mult|. 
Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes would
be awkward due to problems with equality. We avoided these problems by shifting the fields
from the dependent record to its argument, thus mimicking manifest fields by using type classes.
Since |op| is \emph{instantiated} with |mult|, we need not worry about the
equality |op=mult|. This seems to be a general phenomenon. For instance, instead of searching
for a |Monoid with op=mult| we would use type class resolution to find an instance of |Monoid mult|.
We have thus addressed the problem of multiple inheritance.

\section{Universal algebra}\label{univ}
After having implemented concrete algebraic structures, we generalize some of our constructions
by implementing many-sorted universal algebra. A formalization of universal algebra
already exists in Coq~\cite{DBLP:conf/tphol/Capretta99} and has recently been extended with
more categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of
this using type classes. We define equational theories and their varieties of algebras, i.e.\ their
finitary algebraic category of models. A quote function helps to translate a statement for an
instance of
a type class to a statement in the equational logic. We will come back to this in
Section~\ref{quote}.

% \subsection{Universal algebra continued}\label{univ2}

We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
theorem.~\marginpar{Description of the work.}
Discuss: the duplication of code which may have been avoided by universe
polymorphism for |Fixpoint|s.
%perhaps the soundness theorem for equational logic?

In order to prove the first homophism theorem, we define a congruence.
The compatibility of the relation |e| can be stated in two provably  equivalent ways.
Directly, |eAlgebra := @Algebra et v e _| or  indirectly, by stating that the relation
as a set of pairs is a subalgebra in the product algebra.
%theory/ua_congruence.v
% \begin{code}
%  eSub := @ua_subalgebraT.ClosedSubset et vv _ _ Q.
% \end{code}
Subsequently, we choose for former as a definition:
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
A congruence relation is coarser than the setoid relation, as indicated by the use of |Proper|
above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
congruences and quotients benefits from this.

We define the forgetful functor from the category of $\tau$-algebras to the category of sets.
We define the term algebra consisting of closed terms, terms with the empty type as set of
variables. The term algebra is the free $\tau$-algebra and hence\marginpar{?}
initial. Categorically: we have defined a left-adjoint to the forgetful functor. A formal proof
connects the categorical and the algebraic definition.
One may expect that this term algebra could be useful to
\emph{define}, say, the natural numbers as the term algebra for the theory of semiring. However, it
seems difficult to prove decidability of equality, which requires a normalization procedure.

The product of two $\tau$-algebras is their categorical product.

We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra and an
object in the category of $\\tau$-algebras. Conversely, given an object in the category/variety, we
construct a monoid.
%varieties/monoid.v
\begin{code}
Variable o: variety.Object theory.
Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
Global Instance: MonoidUnit (o tt) := algebra_op theory one.
Global Instance from_object: Monoid (o tt).
\end{code}
In the last line, instance resolution `automatically' finds the operation and the unit specified in
the lines before.\marginpar{Why this way?}

\subsection{Category theory}\label{cats}
Using type classes, we provide a basic library for category theory: categories, functors,
adjunctions, monads,\ldots We follow the usual type theoretical definition of a
category~\cite{saibi1995constructive}

%format ◎ = "\circ"
%interfaces/abstract_algebra.v
\begin{code}
Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
  `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z,
      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
  ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{code}
Where |%signature|
tells Coq how to interpret |equiv|.\marginpar{Where is this defined?}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

%theory/categories.v
Natural transformations are formalized as:
\begin{code}
  Context `{Category C} `{Category D}
  `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.

  Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
    natural: Π (x y: C) (f: x ⟶ y), η y ◎ fmap F f = fmap G f ◎ η x.
\end{code}

Here |fmap F| explicitly refers to the arrow part of the functor |F|.
In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
not seem to fit with type theory.\marginpar{Explain why.}

@@

We have defined the category of setoids, the category of categories, the dual of a category,
products, and hence co-products, the category of $\tau$-algebras.

We provide the construction of the term-algebra for a type $\tau$ and show that it is
initial in the category of $\tau$-algebras. We proceed to prove that this provides a left-adjoint
to the forgetful functor. The resulting expression monad will be used in Section~\ref{quote} below.

\section{Interfaces using category theory}\label{interfaces}\label{modul}
The current work was stimulated by our desire~\cite{Riemann} to move between several
implementations in a flexible way. In haskell type class polymorphism is used for this purpose.
Hence, we do the same in type theory.

We have characterized the naturals as the initial object in the category of semirings and derived
many of their properties from this interface. Unary, binary and machine numbers are
instances of this interface, so we can directly apply these results to all these instances.
Similarly, the integers are the initial ring. The rationals are the field of fractions of the
integers. 
Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra on a set $X$.
We provide two implementations of polynomials: the
standard representation using lists of coefficients and the Bernstein representation. A Bernstein
basis polynomial is one of the form:
\[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.

@@

%There is a slight problem when defining the forgetful functor from the category of
%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
%categories\footnote{Perhaps we should have used higher categories for this.}.
%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
%definitions.

@@

We define a term language for monoids:
\begin{code}
Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
\end{code}
This defines a monad on the category of |Type|s with extensional functions.\marginpar{Proof this
formally}.
The expression type is parameterized over the set of variable indices. Hence, we diverge
 from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
 dummy variables.

\noindent An expression is only meaningful in the context of a variable
assignment:\footnote{Categorically: |nat| is
an |Expr|-algebra and |eval| is |map vars| composed with the algebra map.}
\begin{code}
Definition Value := nat.
Definition Vars V := V -> Value.

Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
  match e with
  | Zero => 0
  | Mult a b => eval vs a * eval vs b
  | Var v => vs v
  end.
\end{code}
%
Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
|Expr|-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
corresponding bind operation, the arrows in the Kleisli category.

@@

% We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
% with extensional functions between them~\marginpar{Really?}.

\noindent We define some simple combinators for variable packs:
%
\begin{code}
Definition novars: Vars False := False_rect _.
Definition singlevar (x: Value): Vars unit := fun _ => x.
Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
  fun i => if i then a j else b j.
\end{code}
